Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics

### 2014-10-02 anderson-etal-blackswan-timeseries.tex
Bankruptcy commit

\documentclass[11pt]{article}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
%\usepackage{pdfpages}
%\usepackage{lscape}
%\usepackage{pdflscape}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{lineno}

\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{ragged2e}
\usepackage{tabu} % for spacing between rows in longtable
\setlength{\tabulinesep}{9pt}

% Palatino
\usepackage[full]{textcomp}
     \usepackage[osf]{newpxtext} % osf for text, not math
     \usepackage{cabin} % sans serif
     \usepackage[varqu,varl]{inconsolata} % sans serif typewriter
     \usepackage[bigdelims,vvarbb]{newpxmath} % bb from STIX
     \usepackage[cal=boondoxo]{mathalfa} % mathcal

 % Linux Libertine:
%  \usepackage{textcomp}
%     \usepackage[sb]{libertine}
%     \usepackage[varqu,varl]{inconsolata}% sans serif typewriter
%     \usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
%     \usepackage[cal=boondoxo]{mathalfa} % mathcal
%     \useosf % osf for text, not math
%     \usepackage[supstfm=libertinesups,%
%       supscaled=1.2,%
%       raised=-.13em]{superiors}

@@

\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series}
\author{
Sean C. Anderson$^{1\ast}$ \and
Trevor A. Branch$^2$ \and
Andrew B. Cooper$^3$ \and
Nicholas K. Dulvy$^1$
}
\date{}

@@

\noindent
\textsuperscript{1}Earth to Ocean Research Group, Department of Biological Sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada

\noindent
\textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA

@@

\section{Abstract}
- introduce black swans and heavy tails, perhaps use the climate and increasing climate variability hook

- we find evidence of heavy tails/black swans in ecological time series, but they are rare

- taxonomically widespread although most frequent in birds, followed by mammals, insects

- we don't find compelling intrinsic biological (e.g.\ lifehistory) correlates

- common causes were extrinsic extreme events or multiple extreme events in a row: climate and severe winters, cycles of natural enemies (parasites and predators) and interactions between these

- also mention the advance of our method: probabilistic in quantifying heavy tailedness, incorporates population dynamics, can allow for autocorrelation and observation error, and incorporate prior information

- conclude with ecological, modelling, policy, and management implications

@@

Increasing realization about the importance of extreme events in the environment (REFs)
and of the importance of ecological surprises and black swans (REFs)

Papers to work in:

@@

The data: briefly introduce the GPDD and how we filtered it. Expand in the supplement. Mention the breakdown of taxonomy and interpolation (Table~S1).

The Student-t Gompertz model is:
\begin{align*}
x_t &= \ln N_t\\
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma),
\end{align*}

\noindent where $x_t$ is the $\ln$ abundance at time $t$. The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependence if $b < 0$. The parameter $\lambda$ represents the expected population growth rate at $x_t = 0$. The process noise, $\epsilon_t$, is modelled as a Student-t distribution that is centered at $0$, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$. If $\nu$ is small ($\lesssim 10$) the distribution has much heavier tails than a normal distribution.  For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}). By estimating the value of $\nu$ we can quantity the degree of heavy-tailedness in the process noise jumps from $\ln N_t$ to $\ln N_{t+1}$.

@@

Mention that we simulation tested the model.

Mention the priors (Fig.~\ref{fig:priors}) and expand on their justification in the supplement.

$\nu$ prior: \citep{fernandez1998}
$\sigma$ prior: \citep{gelman2006c}
consider alternative inverse prior on $\nu$ as in \citep{gelman2014}

mention data used for life-history hypothesis comparisons from \citet{brook2006a}

justify why we only present these as hypotheses but do not formally model them

@@

- quantify how frequent and strong heavy tails were (Fig.~\ref{fig:nu-coefs})

- plot showing all p($\nu < 10$) $> 0.5$ (Fig.~\ref{fig:heavy-ts})

- across how many different species

- break down by taxonomy  (Fig.~\ref{fig:nu-coefs} inset panels)

- no obvious intrinsic biological drivers (Fig.~\ref{fig:correlates})

- mention the common causes (Table~\ref{tab:sparks})

- touch on the supplemental figures --- simulation results (Fig.~\ref{fig:sim-nu}, Fig.~\ref{fig:sim-gompertz}, Fig.~\ref{fig:sim-prob})

- alternative models: (Fig.~\ref{fig:ar1}, Fig.~\ref{fig:gomp-vs-ricker})

@@

- why we think we see them in some taxa but not others (observation scale dynamics mismatch; observation errors...)

Ways forward:

- hierarchical modelling of nu and other parameters

- better datasets for specific taxa (e.g.\ fish recruitment)

- joint prior on scale and nu parameters

- formally investigating what heavy tails are associated with

- do heavy tails have conservation importance?

- can we forecast their probability and develop correlates?

- how can we detect them relatively quickly after they happen?

- how often are heavy tails real? observation error, recording errors...

- what does this mean for how we model time series (should we consider using the t as a default even if we fix nu?)

- what does this mean for conservation management (crashes might be more frequent and greater in magnitude than our typical models tell us) --- importance of establishing polices that are robust to this

@@

@@

GPDD

Funding

@@

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.46\textwidth]{nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes. Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations. Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order. Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
Taxonomic orders with three or fewer populations in panel a are omitted for space.}\label{fig:nu-coefs}
\end{center}
\end{figure}

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=\textwidth]{gomp-base-p10-dot-order.pdf}
%\caption{
%  Probability $\nu<20$, i.e.\ one way of asking what the probability is of heavy tails.
%  Dots represent populations.
%  Something like this could be made more pretty and sorted in some appropriate way\ldots maybe with little silhouettes for the animals.
%  Looks pretty similar for $\nu < 10$, with the probabilities a bit lower.
%  There is some vertical jitter added. Could do this and have another subpanel that breaks down the mammals by family, since there's the most interesting stuff happening there and there are lots of them... or just colour the dots by family within each panel.}
%\label{fig:dots}
%\end{center}
%\end{figure}

@@

Using Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014} I'm starting with 4 chains and 1000 iterations with the first 500 as warmup (i.e.\ 2000 total samples). If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again. These conditions were met by 8000 iterations (16000 total samples) in all cases and in almost all cases (X\%) within the initial 1000 iteration case (i.e.\ a total of 2000 samples).

@@

We also fitted a Ricker-logistic model:
\begin{align*}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma),
\end{align*}
\noindent
where  $r_\mathrm{max}$ represents the maximum population growth rate that is obtained when $N$ (abundance) $= 0$. The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at time $t$. The Ricker-logistic model assumes a linear decrease in population growth rate ($x_t / x_{t-1}$) with increases in abundance ($N_t$). In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in $\ln$ abundance ($x_t$) (REF).

To fit the Ricker-logistic models ,we chose a prior on $K$ uniform between 0 and twice the maximum observed abundance (\citet{clark2010} chose 0 and maximum observed, which is less conservative). We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}. We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.

@@

We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma).
\end{align*}
\noindent
In addition to the parameters in the original Gompertz model, an additional parameter, $\phi$, is estimated that represents the relationship between of subsequent process noise residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = -0.99, \mathrm{max.} = 0.99)$.

@@

\noindent
where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at $0.3$. TODO justify $0.3$.

@@

% \begin{figure}[htbp]
% \begin{center}
% \includegraphics[width=\textwidth]{nu-covariates-p-base.pdf}
% \caption{Same as previous figure but the y axis is p(nu $<$ 20), which I think might be better to use\ldots, in reality this kind of coarse dichotomy I think is more realistic. If p(nu $<$ 20) is high (say above 0.5) then there's a good chance the tails are heavy. Otherwise, we just don't know.}
% \label{default}
% \end{center}
% \end{figure}

%\begin{figure}[htbp]
%\begin{center}
%\includegraphics[width=0.7\textwidth]{p10-mammals-cross.pdf}
%\caption{Probability that nu $<$ 10 vs.\ log10(body mass) and log10(years at sexual maturity).}
%\label{default}
%\end{center}
%\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{ts-gpdd-heavy-eg-log10-no-fur.pdf}
\caption{Time series for populations with p($\nu < 10$) $>$ 0.5 using the base Gompertz population model. Panels are ordered by increasing p($\nu < 10$). Vertical axes are on a log10 scale. Colours indicate taxonomic classes. TODO add dots for upswings and downswings as in Table 1. The labels on each panel indicate p($\nu < 10$) $>$ 0.5, the common name for the species, and the GPDD ID number.}
\label{fig:heavy-ts}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.9\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process error (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows). The red lines indicate the true population value. When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples. TODO switch to nu = black swan, 3, 5, and normal.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=1.1\textwidth]{sim-gompertz.pdf}
\caption{Simulation testing the Gompertz estimation model when the process error is chosen so that $\nu$ can be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.5 of specified $\nu$).
  The simulation was run across population $\nu$ values (columns) and different scenarios (colours): (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.3^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
  Dashed horizontal lines show the true population values;
  these true values were chosen to represent approximately the median values as estimated from the GPDD.
  Individual dots and lines represent a stochastic draws from the true population distribution and a model fitting.
  The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and not heavy tailed).
  The panels from top to bottom show $1/\nu$, process noise scale parameter $\sigma_\mathrm{proc}$, growth rate parameter $\lambda$, and the density dependence parameter $b$.
  Dots and line segments are medians and 80\% credible intervals of the posterior.}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{check-sim-box.pdf}
\caption{The same simulations results as the previous figure but here represented as probability that $\nu < 10$ (i.e.\ the probability of heavy tails). Within each scenario the dots represent stochastic draws from the true population distributions.}
\label{fig:sim-prob}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.9\textwidth]{gomp-vs-logistic-nu.pdf}
\caption{Gompertz vs.\ Ricker logistic $\nu$ (nu) estimates. Line segments show 50\% credible intervals; dots show median estimates. The vast majority of heavy-tailed estimates are similar for the two population models. In a few instances one model estimates heavy tails where the other does not. This happens more often for the Ricker-logistic than the Gompertz model.}
\label{fig:gomp-vs-ricker}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{effect-of-ar1-on-nu.pdf}
\caption{The effect of assuming no autocorrelation of residuals (vertical axis) vs.\ estimating autocorrelation of residuals (horizontal axis). Shown are medians and 50\% credible intervals of the posterior. The colours indicate $\phi$ --- the parameter estimating the lag-1 correlation of the residuals. The diagonal line illustrates a one-to-one relationship. There are a few populations in which the heavy-tail estimate changes when autocorrelation is incorporated, but a systematic difference.}
\label{fig:ar1}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
  density of the Bayesian priors for the base Gompertz model. Per capita growth
  rate at log(abundance) = 0: $\lambda \sim \mathrm{Normal}(0, 10^2)$; scale
  parameter of t-distribution process error: $\sigma_\mathrm{proc} \sim \mathrm{Half\mhyphen
    Cauchy} (5)$; t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Exponential}(0.01)$. Not shown is $b$, the density dependence
  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
  \label{fig:priors}
\end{center}
\end{figure}

@@

\end{document}

1 - there is evidence of heavy tails/black swans... in ecological time series... but they are rare

- taxonomically widespread
- we don't find compelling... intrinsic obvious biological reason
- commonly extrinsic extreme events and interactions - climate and severe winters, cycles of natural enemies (parasites + predators),
- as climate gets more variable, extremes more frequent and greater magnitude (and skewed), might expect more of these
- and as our time series get longer and censusing gets more accurate... more ability to detect these
- calls for policy that is be robust to the possibility of black swans
- lead with ecological application

grouse are hunted - shags have erupted

2 - method
- probabilistic
- you can get away with 'these kinds of mistakes if you allow for heavy-tailed process errors'
-  first with a population dynamics model
- para on why novel

hook - to what degree to extreme environmental events..
population dynamics can often be connected to
will be increase in...
and then, however it remains unclear the degree to which animal populations also exhibit (too neutral, something like suffer) from extreme events...

ecology letters

- argue for why we do what we're doing

some we were able to track down

- digg into Barbary macaque

decreasing median - order plot

4 rows one column

add little years on spark lines at beginning and end at bottom?

switch to nu = 20 (or 10?) in first figure
l
combine fig 2 and 3 into new fig 2

new fig3 is cross plots
b, lambda, sigma_proc, N,
log10length, log10Lifespan or min age at maturity,

add 'black swan' simulation in which single value chosen at deviation = 8 or something like that

new supp fig that shows cross plot of accounting or not for observation error

show boxes of misclassification in cross plot figure of various model estimates

line up black swans, say with birds in UK - NAO

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

citation with good justification of gompertz and other references: \citep{herrandoprez2014}

The Gompertz was by far the most favoured model in \citet{brook2006} with the GPDD.

I could see an argument being made that we're just seeing them because of autocorrelation in the residuals. Autocorrelation plays a big role in \citet{inchausti2002} justifying why they think they're seeing increasing CV of time series in the GPDD with time and an increasing CV could create heavy tails. They just model the spectral properties and CV of the abundance time series --- not a population dynamics model. I tried estimating a coefficient for the first-order autocorrelation of the residuals. There is some autocorrelation it picks up sometimes, but interestingly this doesn't change things that much with respect to heavy tail results. Nothing systematic at least (Another important reference on spectral analysis and the GPDD: \citet{garcia-carreras2011}.)

little evidence for benefit of including annual climate or climate extremes with these data:
\citep{gregory2010}
and little evidence for allee effect with them

the natural logorithm version of the logistic is often convenient for parameter estimation
\citep{valpine2002} (and they cite a couple people too)

cite Limited evidence for the demographic Allee effect from numerous species across taxa
Stephen D. Gregory1,4, Corey J. A. Bradshaw2,3, Barry W. Brook2, and Franck Courchamp1
Read More: http://www.esajournals.org.proxy.lib.sfu.ca/doi/full/10.1890/09-1128.1

###

\item illustration of how the exponential prior slightly constrains the sampling of $\nu$\ldots illustrating what the prior probability of heavy tails is given uninformative data (about 5\% probability $\nu < 10$.

\item plots of the priors overlayed with estimates across populations for all parameters

\item cross plot of $\nu$ estimates from the models: no density dependence, add density dependence, add AR1, add assumed observation error

\item time series plots of all heavy-tailed populations

>>>>

\documentclass[12pt]{article}
\usepackage{geometry}
\geometry{verbose,letterpaper,tmargin=2.54cm,bmargin=2.54cm,lmargin=2.54cm,rmargin=2.54cm}
\geometry{letterpaper}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{setspace}
\usepackage{booktabs}
\usepackage{lineno}
\usepackage{ragged2e}

\usepackage{longtable}
\usepackage{multirow}
\usepackage{array}
\usepackage{tabu} % for spacing between rows in longtable
\setlength{\tabulinesep}{9pt}

 % Linux Libertine:
  \usepackage{textcomp}
     \usepackage[sb]{libertine}
     \usepackage[varqu,varl]{inconsolata}% sans serif typewriter
     \usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
     \usepackage[cal=boondoxo]{mathalfa} % mathcal
     \useosf % osf for text, not math
     \usepackage[supstfm=libertinesups,%
       supscaled=1.2,%
       raised=-.13em]{superiors}

@@

\title{Black swans in ecological time series\\Black swans in population dynamics\\Heavy-tailed ecological time series\\Black swans in ecological dynamics}
\author{
Sean C. Anderson$^{1\ast}$ \and
Trevor A. Branch$^2$ \and
Andrew B. Cooper$^3$ \and
Nicholas K. Dulvy$^1$
}
\date{}

@@

\noindent
\textsuperscript{1}Earth to ocean research group, department of biological sciences, Simon Fraser University, Burnaby BC, V5A 1S6, Canada
\noindent
\textsuperscript{2}School of Aquatic and Fishery Sciences, University of Washington, Box 355020, Seattle, WA 98195, USA

@@

\noindent
\textit{Main message, to be deleted (25 words)}: Black swans are present \& taxonomically widespread but rare in population dynamics.
Extreme climate, predation, parasites, and their interactions are common causes; intrinsic drivers are unclear.

%When they do occur they tend to be driven by ... little evidence....

\section{Abstract}

Black swans are statistically improbable events that have profound implications when they occur. Such events are known to occur in financial, social, and environmental systems; the magnitude of environmental extremes is projected to increase with climate change. But, the prevalence and magnitude of black swans in ecological population dynamics is unknown. Here, we develop a probability model that allows us to estimate the degree of heavy-tailedness in ecological process noise. We apply our model to \NPops\ time series from around the world across \NOrders\ taxonomic orders and seven classes. We find strong evidence of black swans, but they are rare, occurring in \overallMinPerc--\overallMaxPerc\% of populations and most frequently for birds (\AvesRangePerc\%) followed by mammals (\MammaliaRangePerc\%), insects (\InsectaRangePerc\%), and fishes (\OsteichthyesRangePerc\%). When they occur, they tend to be driven by climate and severe winters, cycles of natural enemies (parasites and predators) and interactions between these elements. We find little evidence of intrinsic life-history correlates.
Our results suggest [FOR ECOLOGICAL MODELLING]
and the importance of establishing management strategies that are robust to extreme events \ldots.

%lso mention the advance of our method: probabilistic in quantifying heavy tailedness, incorporates population dynamics, can allow for autocorrelation and observation error, and incorporate prior information

%- conclude with ecological, modelling, policy, and management implications

@@

\begin{enumerate}
\item Black swans are statistically improbable events whose occurrence has major ramifications. One of the most profound black swans in ecology was the asteroid marking the mass extinctions at the K-T boundary. Today, it is the extremeness of climate that is expected to cause the greatest societal damage. While extremes in the physical environment are present and widely understood, it remains unclear the extent to which ecological systems buffer or suffer from black swans.

\item What is the existing evidence for heavy tails in ecological time series?

\item There are two main possible reasons for why we may find little evidence of ecological black swans. (1) don't exist, (2) a paucity of appropriate methods or length/quality of time series.

\item Here we develop \ldots our research questions and main conclusions.

%Increasing realization about the importance of extreme events in the environment (REFs)
%and of the importance of ecological surprises and black swans (REFs)

\bigskip
Papers to work in:

@@

\begin{enumerate}
  \item Short overview paragraph
  \item The data: global population dynamics database; briefly introduce the GPDD and how we filtered it. Expand in the supplement. Mention the breakdown of taxonomy and interpolation (Table~S1).
  \item The heavy-tailed Gompertz population model. Mention the simulation testing but don't go into details here.
  \item Mention the alternative population models we fit.
  \item Modelling the possible correlates (mixed effects models with binomial or beta distributions and logit links; nested random effects for taxonomic class and taxonomic order). Life-history data from \citet{brook2006a}.
  \end{enumerate}

The Student-t Gompertz model is:
\begin{align*}
x_t &= \ln N_t\\
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}(\nu, 0, \sigma),
\end{align*}

\noindent where $x_t$ is the $\ln$ abundance at time $t$. The model is density independent if $b = 1$, maximally density dependent if $b = 0$, and inversely density dependence if $b < 0$. The parameter $\lambda$ represents the expected population growth rate at $x_t = 0$. The process noise, $\epsilon_t$, is modelled as a Student-t distribution that is centered at $0$, has a scale parameter of $\sigma_\mathrm{proc}$, and has a degrees of freedom parameter of $\nu$. If $\nu$ is small ($\lesssim 10$) the distribution has much heavier tails than a normal distribution.  For example, at $\nu = 2$, the probability of drawing a value less than -5 is 1.8\%, whereas the probability of drawing such a value from a normal distribution is nearly zero ($2.9\cdot10^{-5}$\%). As $\nu$ approaches infinity the distribution approaches the normal distribution (Fig~\ref{fig:didactic}). By estimating the value of $\nu$ we can quantity how heavy-tailed the process noise deviations are.

@@

%Mention that we simulation tested the model.

%Mention the priors (Fig.~\ref{fig:priors}) and expand on their justification in the supplement.

%$\nu$ prior: \citep{fernandez1998}
%$\sigma$ prior: \citep{gelman2006c}
%consider alternative inverse prior on $\nu$ as in \citep{gelman2014}

%justify why we only present these as hypotheses but do not formally model them

\bigskip
Supporting material to be noted:

  \item Table showing the taxonomic breakdown, number of interpolated points, etc (Table~S1).

  \item Plot showing time series we assumed were recorded as log transformed  (Fig.~\ref{fig:log10-assumed}).

  \item Plot showing all the time series we used with interpolated and zero imputed values (Fig.~\ref{fig:all-ts}).

  \item Plot showing the priors (Fig.~\ref{fig:priors}).
\end{enumerate}

@@

Main figures and tables:

\begin{enumerate}
  \item Fig.~\ref{fig:didactic} illustrates the method: it shows the t-distribution tails, example simulated time series, and model fits to those time series.

  \item Fig.~\ref{fig:nu-coefs} shows the posterior distributions of the estimated $\nu$ values. These are split by taxonomic class and order.

  \item Fig.~\ref{fig:correlates} shows possible biological and time-series-property correlates of heavy-tailed behaviour.

  \item Table.~\ref{tab:sparks} selects populations that were categorized as heavy-tailed and digs into the causes. These are a non-random sample that I was able to verify in the literature.

%- quantify how frequent and strong heavy tails were (Fig.~\ref{fig:nu-coefs})

Supporting material to be noted quickly:

  \item Simulation testing: the ability to recover $\nu$ when randomly sampling from various distributions (Fig.~\ref{fig:sim-nu}); heavy-tailed Gompertz model performance and confidence interval coverage given the process noise has deviations that are known to have effective $\nu$ equal to true $\nu$ (Fig.~\ref{fig:sim-gompertz}); boxplots of the same output (Fig.~\ref{fig:sim-gompertz-boxplots}); probability that $\nu < 10$ for the Gompertz simulation testing (Fig.~\ref{fig:sim-prob}).

  \item Time series of all heavy tailed populations (Fig.~\ref{fig:heavy-ts}).

  \item The effect of alternative population dynamics models on $\nu$ estimates (Fig.~\ref{fig:alt}).

  \item Coefficients from modelling covariates of the probability of heavy tails (Fig.~\ref{fig:correlate-coefs}).

  \item Example Stan code for heavy-tailed Gompertz model with AR1 residuals and observation error (a model including everything).

  \item GPDD IDs used.

Modelling result: an increase of 1 time step of data (given that you are ballpark around the mean --- 30 time steps --- to start with) results in an approximately 1\% increase in the expected probability of observing heavy tails. (Using Gelman's `divide by 4' rule for interpreting logistic regression coefficients.)

@@

\begin{enumerate}
  \item Summary of our findings
  \item How do our results mesh with previous related analyses?
  \item Why might we expect to see heavy tails in ecological time-series? (mixture of normals or extreme drivers)
  \item Are the observed frequencies by taxonomic class real or an observational phenomena?
  \item Ways forward
  \item Policy implications
\end{enumerate}

@@

Funding: SCA: SFU Graduate Fellowship, NKD: NSERC SFU \ldots, ABC: \ldots, TAB: \ldots

Earth to Ocean's research group for helpful discussions

Global Population Dynamics Database

@@

Compute Canada's WestGrid high-performance computing resources

@@

%\clearpage

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.42\textwidth]{nu-coefs-2.pdf}
\caption{
Estimates of how heavy-tailed population dynamics are for \nuCoefPopN\ populations of birds, mammals, insects, and fishes. Small values of $\nu$ ($\lesssim 10$) suggest heavy-tailed dynamics; larger values of $\nu$ suggest approximately normal-tailed dynamics. Vertical points and line segments represent posterior medians and 50\% / 90\% credible intervals for individual populations. Inset plots show probability that $\nu < 10$ (probability of heavy tails) for populations arranged by taxonomic order. Taxonomic orders are sorted by decreasing mean p($\nu < 10$).
Taxonomic orders with three or fewer populations in panel a are omitted for space.}\label{fig:nu-coefs}
\end{center}
\end{figure}

@@

We fit our models with Stan 2.4.0 \citep{stan-manual2014}, and R 3.1.1 \citep{r2014}. I'm starting with 4 chains and 2000 iterations with the first 1000 as warmup (i.e.\ 4000 total samples). If rhat is greater than 1.05 for any parameter or the minimum effective sample size is less than 200 for any parameter then I double both the total iterations and warmup and run again. This continued up to 8000 iterations (16000 total samples) \ldots

@@

We also fitted a Ricker-logistic model:
\begin{align*}
x_t &= x_{t-1} + r_{\mathrm{max}}\left(1 - \frac{N_{t-1}}{K}\right) + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(0, \sigma),
\end{align*}
\noindent
where  $r_\mathrm{max}$ represents the maximum population growth rate that is obtained when $N$ (abundance) $= 0$. The parameter $K$ represents the carrying capacity and, as before, $x_t$ represents the $\ln$ transformed abundance at time $t$. The Ricker-logistic model assumes a linear decrease in population growth rate ($x_t - x_{t-1}$) with increases in abundance ($N_t$). In contrast, the Gompertz model assumes a linear decrease in population growth rate with increases in $\ln$ abundance ($x_t$) (REF).

To fit the Ricker-logistic models, we chose a prior on $K$ uniform between zero and twice the maximum observed abundance (\citet{clark2010} chose uniform between zero and maximum observed, which is less conservative). We set the prior on $r_\mathrm{max}$ as uniform between 0 and 20 as in \citet{clark2010}. We used the same priors on $\nu$ and $\sigma$ as in the Gompertz model.

@@

We considered a version of the Gompertz model in which an autoregressive parameter was fit to the process noise residuals:
\begin{align*}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student\mhyphen t}_\nu(\phi \epsilon_{t-1}, \sigma).
\end{align*}
\noindent
In addition to the parameters in the original Gompertz model, an additional parameter, $\phi$, is estimated that represents the relationship between of subsequent process noise residuals. Based on the results of previous analyses with the GPDD \citep[e.g.][]{connors2014} and the chosen priors in previous analyses \citep[e.g.][]{thorson2014a} and to greatly speed up chain convergence when running our model across all populations, we placed a weakly informative prior on $\phi$ that assumed the greatest probability density near zero with the reduced possibility of $\phi$ being near $-1$ or $1$. Specifically, we chose $\phi \sim \mathrm{Truncated\mhyphen Normal}(0, 1, \mathrm{min.} = 1, \mathrm{max.} = 1)$.

@@

\noindent
where $U$ represents the unobserved state vector, and $\sigma_\mathrm{obs}$ represents the standard deviation of observation error (on a log scale), which was set at $0.2$. TODO justify $0.2$.

@@

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{priors-gomp-base.pdf} \caption{Probability
  density of the Bayesian priors for the base Gompertz model. From left to right: (1) per capita growth
  rate at $\ln$(abundance) = $0$: $\lambda \sim \mathrm{Normal}(0, 10^2)$; (2) scale
  parameter of t-distribution process noise: $\sigma \sim \mathrm{Half\mhyphen
    Cauchy} (2.5)$; (3) t-distribution degrees of freedom parameter: $\nu \sim
  \mathrm{Exponential}(0.01)$. (4) AR1 correlation coefficient of residuals: $\phi \sim \mathrm{Normal}(0, 1)[-1, 1]$. Not shown is $b$, the density dependence
  parameter: $b \sim \mathrm{Uniform}(-1, 2)$.}
  \label{fig:priors}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-prior-exp0point01.pdf}
\includegraphics[width=0.8\textwidth]{t-dist-sampling-sim-sigma-prior-exp0point01.pdf}
\caption{
  Testing the ability to estimate $\nu$ (top panels) and the scale parameter of the process error (bottom panels) for a given number of samples (columns) drawn from a distribution with a given true $\nu$ value (rows). The red lines indicate the true population value. When a small number of samples are drawn there may not be samples sufficiently far into the tails to recapture the true $\nu$ value; however, heavy tails are still distinguished from normal tails in most cases, even with only 25 or 50 samples. TODO switch to nu = black swan, 3, 5, and normal.
}
\label{fig:sim-nu}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{sim-gompertz.pdf}
\caption{Simulation testing the Gompertz estimation model when the process error draws are chosen so that $\nu$ can be estimated close to the true value outside the full population model (``effective $\nu$'' within a CV of 0.2 of specified $\nu$).
  The simulation was run across population $\nu$ values (columns) and different scenarios (colours): (1) 100 time steps and no observation error, (2) 50 time steps and no observation error, (3) 50 time steps and observation error drawn from $\mathrm{Normal} (0, 0.2^2)$ but ignored, and (4) 50 time steps with observation error in which the quantity of observation error was assumed known.
  Dashed horizontal lines show the true population values;
  these true values were chosen to represent approximately the median values as estimated from the GPDD.
  Individual dots and lines represent a stochastic draws from the true population distribution and a model fitting (median and 80\% credible intervals).
  %The panels show true values of $\nu$ =  3, 5, normal (very heavy, heavy, and not heavy tailed).
  The panels from top to bottom show $1/\nu$, process noise scale parameter $\sigma$, growth rate parameter $\lambda$, and the density dependence parameter $b$.}
\label{fig:sim-gompertz}
\end{center}
\end{figure}

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=\textwidth]{sim-gompertz-boxplots.pdf}
\caption{The distribution of median estimates from the same simulation as described in Fig.~\ref{fig:sim-gompertz}. The boxes show the interquartile range. The whiskers extend to $1.5$ times the interquartile range and outliers are shown as dots.}
\label{fig:sim-gompertz-boxplots}
\end{center}
\end{figure}
